Application of Equal Local Levels to Improve Q-Q Plot Testing Bands with \proglang{R} Package \pkg{qqconf}

### 2021-07-08 article/manuscript.Rnw
updated neuro citation

\subsection{Multiple Testing in Genome Wide Association Studies}
\subsubsection{Applicability of Equal Local Levels to Genome Wide Association Studies}
In genome wide association studies (GWAS), we are interested in identifying the genetic variants that are responsible for biological traits. Often, we test a large number of single-nucleotide polymorphisms (SNPs), to see if any of them are strongly correlated with some discrete or continuous trait. Often, it is useful to test the following global null hypothesis:
\newline
\newline
$H_{0}$: All genome-wide SNPs in some functional category have correlation 0 with the trait of interest.
%
\begin{equation*}
\textrm{ vs. }
\end{equation*}
%
\noindent$H_{A}$: At least one SNP in some functional category has non-zero correlation with the trait of interest.
\newline
\newline
Typically, when testing this global null, we apply some test to each of $n$ SNPs, yielding $n$ p-values. While the SNPs have a local correlation structure, this has only a weak effect at a genome-wide scale. Thus, the local levels procedure still provides a valuable visualization tool to assess the extent and type of deviation from the null, even though it is not exact. Particularly in genetics, if the p-values deviate from the null, it can be very useful to view graphically how they deviate. For instance, if a few tests yield very small p-values but the p-values from the rest of the tests looks relatively uniform, this indicates that the trait of interest is likely driven by a few mutations. If, however, there are some small deviations throughout the p-value distribution, this indicates that trait of interest is likely driven by a number of genes that all play some small part in a complex biological process, or potentially by some confounding variables that are not controlled for.
\newline
\newline
If a very specific distribution for $H_{A}$ is suspected, then it makes sense to use a test that has high power under this alternative, which may not be our local levels test. However, in the general case when little is known about the alternative distribution, using local levels is desirable because it is agnostic to the space of alternative distributions. Moreover, since small p-values are often of great interest in GWAS because they can indicate the main biological drivers of a trait, the local levels test is far superior to the KS test due to its comparative tail sensitivity.
\subsubsection{Appliction of Equal Local Levels to Creutzfeld-Jakob Disease}
We downloaded the p-values from a GWAS of Creutzfeld-Jakob disease in a sample of XXX cases and YYY controls. The GWAS found strong genetic associations on chromosomes A and B. Here, we remove those chromosomes from the results in order to display the p-values from tests that are not obviously from a non-uniform distribution. To more clearly allow for display of the individual p-values, we downsample the number of tests to 10,000 by extracting approximately equally spaced genetic markers.
<<load_data>>=
cjd_df <- read.table("data/cjd_sample.txt", header = TRUE)
@
We then make QQ plots of these 10,000 tests with 95\% testing bands. First, we make a standard QQ plot of observed quantiles versus expected quantiles (Figure 8).
<<cjd_standard_qq, fig=TRUE, include=FALSE>>=
library(qqconf)
qq_conf_plot(
  obs = cjd_df[,3],
  distribution = qunif,
  points_params = list(pch = 21, cex = 0.2)
  )
@

@@

>>>>

\subsection{Multiple Testing in Genome Wide Association Studies}
\subsubsection{Applicability of Equal Local Levels to Genome Wide Association Studies}
In genome wide association studies (GWAS), we are interested in identifying the genetic variants that are responsible for biological traits. Often, we test a large number of single-nucleotide polymorphisms (SNPs), to see if any of them are strongly correlated with some discrete or continuous trait. Often, it is useful to test the following global null hypothesis:
\newline
\newline
$H_{0}$: All genome-wide SNPs in some functional category have correlation 0 with the trait of interest.
%
\begin{equation*}
\textrm{ vs. }
\end{equation*}
%
\noindent$H_{A}$: At least one SNP in some functional category has non-zero correlation with the trait of interest.
\newline
\newline
Typically, when testing this global null, we apply some test to each of $n$ SNPs, yielding $n$ p-values. While the SNPs have a local correlation structure, this has only a weak effect at a genome-wide scale. Thus, the local levels procedure still provides a valuable visualization tool to assess the extent and type of deviation from the null, even though it is not exact. Particularly in genetics, if the p-values deviate from the null, it can be very useful to view graphically how they deviate. For instance, if a few tests yield very small p-values but the p-values from the rest of the tests looks relatively uniform, this indicates that the trait of interest is likely driven by a few mutations. If, however, there are some small deviations throughout the p-value distribution, this indicates that trait of interest is likely driven by a number of genes that all play some small part in a complex biological process, or potentially by some confounding variables that are not controlled for.
\newline
\newline
If a very specific distribution for $H_{A}$ is suspected, then it makes sense to use a test that has high power under this alternative, which may not be our local levels test. However, in the general case when little is known about the alternative distribution, using local levels is desirable because it is agnostic to the space of alternative distributions. Moreover, since small p-values are often of great interest in GWAS because they can indicate the main biological drivers of a trait, the local levels test is far superior to the KS test due to its comparative tail sensitivity.
\subsubsection{Appliction of Equal Local Levels to Creutzfeld-Jakob Disease}
We downloaded the p-values from a GWAS of Creutzfeld-Jakob disease in a sample of 4110 cases and 13,569 controls \citep{jones2020identification}. Test of association between risk for the disease and genetic variants was done at 6,314,492 SNPs. Major genetic risk loci were found on chromosomes 1 and 20. Here, we remove those chromosomes from the results in order to display the p-values from tests that are not obviously from a non-uniform distribution. For display purposes, we subsample the remaining SNPs to 10,000 approximately evenly spaced SNPs. Subsampling helps ensure that correlations between SNPs that are located
near each other is minimized.
<<load_data>>=
cjd_df <- read.table("data/cjd_sample.txt", header = TRUE)
@
We then make QQ plots of these 10,000 tests with 95\% testing bands. First, we make a standard QQ plot of observed quantiles versus expected quantiles (Figure 8).
<<cjd_standard_qq, fig=TRUE, include=FALSE>>=
library(qqconf)
qq_conf_plot(
  obs = cjd_df[,3],
  distribution = qunif,
  points_params = list(pch = 21, cex = 0.2)
  )
@

@@

From the QQ plots, we can see that there is an excess of small p-values, indicating that the test statistics do not follow the expected null distribution.
